---
title: "Part 3; Tools for Reproducible (Open) Science"
author: '[Marc A.T. Teunis, PhD](https://www.hu.nl/onderzoek/onderzoekers/marc-teunis)'
date: "`r Sys.time()`"
output:
  ioslides_presentation:
    highlight: tango
  powerpoint_presentation:  
subtitle: Why, What and How in a series of three webinars
always_allow_html: yes
widescreen: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  error = FALSE
  , 
  fig.width = 6,
  fig.height = 4
)

image_dir <- here::here(
  "images"
)


## packages
library(tidyverse)
library(ggExtra)

```

## Resources

Get these slides at:

 - https://github.com/uashogeschoolutrecht/work_flows" (source code)
 - https://uashogeschoolutrecht.github.io/ (slides only)

## Tools to enable Open Science (1/3)

1 - Data Science programming languages

```{r, dpi = 80}
knitr::include_graphics(
  file.path(
    image_dir,
    "data_science_languages",
    "InkedDia1_LI.jpg"
  )
)
```


## Tools to enable Open Science (2/3)

2 - Data Science infrastructure & software

```{r, dpi = 80}
knitr::include_graphics(
  file.path(
    image_dir,
    "data_science_languages",
    "Dia2.jpg"
  )
)
```

## Tools to enable Open Science (3/3)

3 - Data Science learning tools

```{r, dpi = 80}
knitr::include_graphics(
  file.path(
    image_dir,
    "data_science_languages",
    "Dia3.jpg"
  )
)
```


## Exploring the tools to do (Open) Data Science at HU

 >- A little introduction to data formats and shapes
 >- Introducing the tools; Github + RStudio + HU ResearchDRive
 >- Introducing RStudio and the `{tidyverse}`
 >- Getting materials from Github **LIVE DEMO**
 >- Getting access to HU ResearchDrive **LIVE DEMO**
 >- (Getting access to HU ResearchDrive from within RStudio **LIVE DEMO**)

## Data-formats - Non-Proprietary {.build}

File format source code is open and maintained by open source community or core development team.

  >- `.netCDF` (Geo, proteomics, array-oriented scientific data)  
  >- `.xml` / `.mzXML` (Markup language, human and machine readable, metadata + data together)
  >- `.txt` / `.csv` (flat text file, usually tab, comma or semi colon (`;`) seperated) 
  >- `.json` (text format that is completely language independent)  

_Will remain readable, even if format becomes obsolete_

**When storing a curated dataset for sharing or archiving it is always better to choose a non-proprietary format**

## Data shape
Look at these two tables, what do you notice?
```{r}
table2 %>% head(3)
table3 %>% head(3)
```

<p style="font-size:14px">Both tables are build-in datasets from the {tidyr} package belonging to the [{tidyverse}](https://www.tidyverse.org/) suite of Data Science R packages</p> 

## The `{tidyverse}`

 - Suite of R-packages for Data Science and functional programming
 - https://www.tidyverse.org/
 - Connect to many other tools in R
 
```{r, dpi = 90}
knitr::include_graphics(
  file.path(
    image_dir,
    "tidyverse.png"
  )
)
```
<p style="font-size:14px">From: [Storybench](https://www.storybench.org/getting-started-with-tidyverse-in-r/)</p>  

## Tidy data
```{r}
knitr::include_graphics(
  file.path(image_dir, "tidy-1.png")
)
```

 1. Each variable goes in its own column
 1. Each observation goes in its own row
 1. Each cell contains only one value

<p style="font-size:14px">From: ["R for Data Science", Grolemund and Wickham](https://r4ds.had.co.nz/)</p> 

## Are these dataframes tidy? 
```{r}
map(
  .x = list(
    table5,
    table4a),
  .f = head, 3)
```
_What steps would we need to tidy them?_ 

## `pivot_longer()`
```{r, echo=TRUE}
table4a %>% 
  pivot_longer(
    cols = `1999`:`2000`, 
    names_to = "year",
    values_to = "cases"
  ) -> table4a_tidy
table4a_tidy
```

## `separate()`
```{r, echo=TRUE}
table5 %>%
  separate(
    col = rate,
    into = c("cases", "population"),
    remove = TRUE
  ) -> table5_tidy
table5_tidy
```

## The git/Github.com workflow; segregating data from compute infrastructure from code

```{r, dpi = 70}
knitr::include_graphics(
  file.path(image_dir,
            "git_workflow.jpg")
)
```

## Just for kicks, a graph
```{r, fig.width=8, fig.height=5}

shipman_murders <- read_csv(
  here::here(
    "data",
    "D030",
    "00-1-shipman-confirmed-victims-x.csv"
    )
  ) %>%
  janitor::clean_names()

plot <- shipman_murders %>%
ggplot(aes(
  x = fractional_death_year,
  y = age,
  colour = reorder(gender2, gender)
  )) +      
  geom_point(size = 2) +
  labs(x = "Year", y = "Age of victim") +
  scale_x_continuous(breaks = seq(1975, 1995, 5),
    limits = c(1974, 1998)) +
  scale_y_continuous(breaks = seq(40, 90, 10), limits = c(39, 95)) +
  scale_size_continuous(name = "Size", guide = FALSE) +
  scale_colour_brewer(palette = "Set1")  + 
  theme_bw() +
  theme(
    legend.position = c(0.125, 1.12),
    legend.background = element_rect(colour = "black"),
    legend.title = element_blank()) +
  ggtitle("Shipman murders")

  ggMarginal(plot, type="histogram")

```
<p style="font-size:14px">[Spiegelhalter, 2020, "The Art of Statistics"](https://github.com/dspiegel29/ArtofStatistics)</p> 

## Getting the code for all webinars into your RStudio environment; introducing the jargon {.build}

 >- **Get an Rstudio installation or account (via me)**
 >- **Clone the repo to your RStudio Env.**
 >- **Install any code dependencies in your Env.** 
 >- **Run the code, and adapt if you want**
 >- *Work on the code*
 >- *Create a commit*
 >- *Create a pull request*
 
**LIVE DEMO** 

## Github user-account

 >- https://github.com
 >- You can create personal and private repos
 >- Adding a README.md to each repo is a good idea
 >- The HU Github Data Science repos: https://github.com/uashogeschoolutrecht  
 
## RStudio

>- Integrated Development for R (and Python, Stan, C++, D3, SQL)
>- Favorite IDE for using R
>- Many integrated productivity tools (auto-completion, syntax highlighting, code-formatting, git-integrations, building tools)
>- Send me an email if you want to use R/RStudio yourself!

## Getting Github-repo content in RStudio {.build}

 >- Copied url to Github repo on clipboard
 >- Open new RStudio Project
 >- Choose 'Version Control' Option
 >- Paste url from clipboard in url field
 >- Let the clone finish
 >- Start using the code!
 >- My code will work from a cloned github repo in an RStudio Project because of the [`{here}` package!](https://github.com/jennybc/here_here) 

_Stop using `setwd()`!!!_

 **LIVE DEMO**

## HU ResearchDrive

 >- Service brought to HU by SURF
 >- Application: 
https://bibliotheek.hu.nl/onderzoekers/datamanagement/
 >- Access though webinterfacte and other software
 >- SFTP software [CyberDuck](https://cyberduck.io/) (you need admin rights)
 >- Rclone (commandline interface)

## Which tool for what?
```{r, dpi = 70}
knitr::include_graphics(
  file.path(
    image_dir,
    "Presentatie1",
    "Dia26.JPG"
  )
)
```

## Access HU-ResearchDrive from RStudio
To make this work you will need a WebDav token from HU ResearchDrive 

`Profile -> Security -> Create App`

```{r, dpi = 80}
knitr::include_graphics(
  file.path(
    image_dir,
    "Presentatie1",
    "Dia17.JPG"
  )
)
```
**Live Demo**


## Thank you for your attention!

```{r, dpi = 150}
knitr::include_graphics(
  file.path(
    image_dir,
    "Pepper.png")
)
```

**Ejoy the summer & Sta*R*t lea*R*ning:** 

 https://r4ds.had.co.nz/
 
 https://rstudio.com/resources/webinars/
 
 Find all the slides on: https://uashogeschoolutrecht.github.io/